Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-29 joint_inf.tex
main text

\begin{abstract}
Maximum likelihood estimation of phylogeny requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We derive a set of parameters that generate data under a four-taxon tree and result in this method estimating an incorrect topology in the limit of infinite-length sequences.
For branch length estimation on the correct topology, we outline similar cases where branch lengths are consistently and heavily biased.
\end{abstract}

@@

Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.
But is it statistically consistent?

Statistical consistency means that an estimator gives the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, doing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

@@

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
This parameter thus quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.

@@

The classical approach is to maximize the likelihood marginalized across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
to estimate the tree $\tau$ and branch fidelities $t$.

An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}.
One can then estimate a topology and branch fidelities using this profile likelihood:
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

We can now state our results on inconsistency of joint inference.
All proofs are deferred to the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
    Due to the looseness of the upper and lower bounds, the parameters in the non-shaded region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
As described above, we use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we know $x^*$ and $y^*$, then for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
we show there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\}).
$$
Again, we find a region for which the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of bad internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch, and then a branch length is estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is set to be that of $y^*$.
Here again, the shaded region is guaranteed to give inconsistent estimation, while the non-shaded region may or may not do so.}
\label{fig:bl-inconsistency}
\end{figure}

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be 1 or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
This theorem can be used to show incorrect branch parameter estimates as follows.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then useful $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity 1 or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
Here again, the shaded region is guaranteed to give such poor estimation, while the non-shaded region may or may not do so.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}).
The region in which $\hat{w}$ is numerically estimated to be 1 is an unusual shape containing the analytically-derived region but is not the same as it.
We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be 1.

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We derive a set of parameters that generate data under a four-taxon tree and result in this method estimating an incorrect topology in the limit of infinite-length sequences.
For branch length estimation on the correct topology, we outline similar cases where branch lengths are consistently and heavily biased.
\end{abstract}

@@

Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
In a recent proposal, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
This latter conditional optimization is simpler and more computationally efficient than optimizing the marginal likelihood.
But is it statistically consistent?

An estimator is statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case where ``branch lengths are short and only a minority of sites change on a give branch,'' their work motivates understanding when exactly this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
In the standard configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, performing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

@@

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
This parameter quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.

@@

The classical approach is to maximize the likelihood marginalized across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \sum_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
\end{equation}
to estimate the tree $\tau$ and branch fidelities $t$.

The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

We now state our results on the inconsistency of joint inference.
All proofs are deferred to the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of fidelities for ``perfect'' data generated on the Farris topology (Fig.~\ref{fig:farris-fels-top}) with $w^* = y^*$.
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we know $x$ and $y$ and estimate them to be the truth, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
%das: @EM Joe pointed out that this is in effect another instance of converging to the wrong topology since once the branch length goes to zero we are no longer in a bifurcating tree case. I don't recall if he had more intuition than the above sentence but he seemed to think it made sense. Not sure if you have any thoughts.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Analytically-derived region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
The shaded region shows the area in which the internal branch length is estimated to be 0, i.e.\ the estimated fidelity $\hat{w} = 1$, even though the generating fidelity is $y^*$.
Here again, the shaded region is guaranteed to give inconsistent estimation, while the white region may or may not do so.}
\label{fig:bl-inconsistency}
\end{figure}

@@

The general case is more challenging to analyze and so we obtain weaker bounds.
Here, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Analytically-derived region of branch parameter inconsistency in terms of fidelities with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the pendant branch parameters.
    As in previous plots, the shaded and white regions are loose indications of inconsistency.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be one.

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher objective function.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.
